\documentclass[dvips, 12pt]{article}

% Any percent sign marks a comment to the end of the line

% Every latex document starts with a documentclass declaration like this
% The option dvips allows for graphics, 12pt is the font size, and article
%   is the style


\usepackage{etex}  % must be include as the first package, otherwise an error will appear
\usepackage[margin=1in]{geometry}

\usepackage{float}
\restylefloat{table}

\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{url}
\usepackage{harmony}
\usepackage{musixtex}
\usepackage{subfigure}
\usepackage{caption}
\usepackage{listings}
\usepackage{syntax}
\usepackage{comment}
\usepackage{color}

\lstset{ %
	    basicstyle=\small\ttfamily, 
		language={Haskell},
		stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
		numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
		numberstyle=\tiny\color{black}  % the style that is used for the line-numbers
}
\renewcommand{\syntleft}{\normalfont\itshape}
\renewcommand{\syntright}{}

%\usepackage{biblatex}
%\usepackage{subcaption}

% These are additional packages for "pdflatex", graphics, and to include
% hyperlinks inside a document.


% These force using more of the margins that is the default style


% Everything after this becomes content
% Replace the text between curly brackets with your own

\title{{\Huge \bfseries SMURF Language Reference Manual} \\ \Large \it Serial MUsic Represented as Functions \vspace{0.6cm}}

\author{\normalsize Richard Townsend, Lianne Lairmore, Lindsay Neubauer, Van Bui, Kuangya Zhai
	\\ \small \{rt2515, lel2143, lan2135, vb2363, kz2219\}@columbia.edu \vspace{0.6cm}}

\date{\today \vspace{2cm}}

% You can leave out "date" and it will be added automatically for today
% You can change the "\today" date to any text you like

\begin{document}
\maketitle
\clearpage


% This command causes the title to be created in the document
\tableofcontents

\section{Introduction}
SMURF is a functional language that allows a composer to create serialist music
based on the twelve-tone composition technique. In general, serialism is a musical composition
method where a set of values, chosen through some methodical progress,
generates a sequence of musical elements. SMURF is based on the
functional syntax and semantics set forth by Haskell. The backend of
SMURF generates MIDIs corresponding to the composition defined by the 
user's initial program in SMURF. A program in SMURF, at its top-most level, 
is a series of declarations separated by newline tokens.

\section{Syntax Notation}
The syntax notation used in this manual is as follows. Syntactic 
categories are indicated by \emph{italic} type. Literal words and 
characters used in the SMURF language will be displayed in \texttt{typeset}. 
Alternatives are listed on separate lines. 

Regular expression notations are used to specify grammar patterns in this
manual.  {\it r}$*$ means the pattern {\it r} may appear zero or more times,
{\it r}$+$ means {\it r} may appear one or more times, and {\it r}$?$
means {\it r} may appear once or not at all. {\it r1}$|${\it r2} denotes an option
between two patterns, and {\it r1 r2} denotes {\it r1} followed by
{\it r2}. 


\input{data/lexical}
\input{data/identifiers}
\input{data/expressions}
\input{data/declarations}
\input{data/libraryFuncs}
\clearpage



\end{document}
